---
title: "Making Cool Graphs with GGPLOT 2"
author: "Juan Mejia"
date: "12/29/2018"
output:
  html_document:
    number_sections: yes
  pdf_document: default
graphics: yes
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 5, fig.height = 4, fig.align = "center")
```

# What will we do today?

We will try to understand the basic structure of GGPLOT2 using two datasets (mpg and diamonds)

Workshop Contents:

- Graphing interface
- Aesthetics mappings
- Facets
- Geometric objects 
- Statistical transformations
- Position Adjustments
- Coordinate Systems
- The layered grammar of graphics

Before getting into the fun stuff there are two things that we must do: 

1. Install Tydiverse (which contains GGPLOT2)
2. Explore the first dataset that we will use (mpg) to understand the dataset nature and the variables involved in our analysis

# Installing Tydiverse, which includes GGPLOT2!

```{r}
 tinytex::install_tinytex(force=TRUE)
```


```{r}
install.packages("mosaic")
install.packages("oilabs")
install.packages("tidyverse", dependencies=TRUE)
library(tidyverse)
```

#Exporing the MPG dataset

```{r}
head(mpg)
```

```{r}
str(mpg)
```


```{r}
?mpg
```


# Basic Graph Example

The graph below uses the data set mpg (pre-loaded in base r) to graph engine size (displ) vs fuel efficiency (hwy)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

# A graphing template

You can think of the structure below as the basic template to use ggplot2 to create awesome graphs. You first havce to chose the data set. Later on you use the GEOM_FUNCTION to select the type of graph (i.e. geometric object) and finally the mappings setting to choose which variables to graphs. 

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

# Aesthetic Mappings (check out the aes parameter in the equation above)

In addition to the basic mapping, we can also use aesthetics to make graphs more insightful. An aesthetic is a visual property of the objects in your plot. Aesthetics include things like the size, the shape, or the color of your points.

In the below graph, we use an aesthetic to show the class of the car (in this dataset, this is the type of car - compact, midsize, suv, etc)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

In the above example, we mapped class to the color aesthetic, but we could have mapped class to the size aesthetic in the same way.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Other aesthetic examples - this time with transparency and point shape: 

```{r}
# Left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Right
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Once you map an aesthetic, ggplot2 takes care of the rest. It selects a reasonable scale to use with the aesthetic, and it constructs a legend that explains the mapping between levels and values. For x and y aesthetics, ggplot2 does not create a legend, but it creates an axis line with tick marks and a label.

You can also set the aesthetic properties of your geom manually. For example, we can make all of the points in our plot blue:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

# Facets

You can use facets within GGPLOT2 to create subplots that display different subests of the data.

To facet your plot by a single variable, use facet_wrap(). The first argument of facet_wrap() should be a formula, which you create with ~ followed by a variable name.

The variable that you pass to facet_wrap() should be discrete.

```{r}
ggplot(data = mpg)+
  geom_point(mapping = aes(x = displ, y = hwy),color = "blue") + 
  facet_wrap(~ class, nrow = 2)
```

The example above shows faceting using only one variable (class). Engine size and fuel efficiency are the x and y axis, not the faceting variable. 

To facet your plot on the combination of two variables, add facet_grid() to your plot call. The first argument of facet_grid() is also a formula. This time the formula should contain two variable names separated by a ~.

In the example below we use the variables drv (front-wheel drive, rear wheel drive or 4wd) and cyl (number of cylinders) to facet our data:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid (drv ~ cyl)
```

# Geometric Objects

A geom is the geometrical object that a plot uses to represent data. People often describe plots by the type of geom that the plot uses. For example, bar charts use bar geoms, line charts use line geoms, boxplots use boxplot geoms, and so on.

Check out the two graphs below: 

```{r}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

The two graphs above graph the same two variables, using the same dataset, but use two different geoms. 
First one is a scatter plot and the second one is a smooth conditional means chart, which uses a logistical regression model, among other methods, to fit the data and smooth it. For more information on geom_smooth, check ot: https://ggplot2.tidyverse.org/reference/geom_smooth.html

Aesthetics in different Geometric Objects:

Every geom function in ggplot2 takes a mapping argument. However, not every aesthetic works with every geom. You could set the shape of a point, but you couldn't set the "shape" of a line. 

On the other hand, you could set the linetype of a line. geom_smooth() will draw a different line, with a different linetype, for each unique value of the variable that you map to linetype.

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color = drv))
  
```

Here geom_smooth() separates the cars into three lines based on their drv value, which describes a car's drivetrain. One line describes all of the points with a 4 value, one line describes all of the points with an f value, and one line describes all of the points with an r value. 

Here, 4 stands for four-wheel drive, f for front-wheel drive, and r for rear-wheel drive.

If this sounds strange, we can make it more clear by overlaying the lines on top of the raw data and then coloring everything according to drv.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))+
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color = drv))
  
```

How can I know which geometric objects (types of plots)I can use?

ggplot2 provides over 30 geoms, and extension packages provide even more (see https://www.ggplot2-exts.org for a sampling). The best way to get a comprehensive overview is the ggplot2 cheatsheet, which you can find at http://rstudio.com/cheatsheets. To learn more about any single geom, use help: ?geom_smooth.

Other Geom_Smooth examples

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
              
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
    
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv)
  )
```

To display multiple geoms in the same plot, in a more succinct and efficient way than the one I used before, you can use the following structure:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Everything included in the GGPLOT parenthesis applies to all the geoms. Everything that is included in a geom mapping applies only to that geometric object. In the graph above, color is used to add an aesthetic element to geom_point, and not to geom_smooth. 

Now, let's look at this code and analyze it together (it is different to the charts we have seen before in that the different mappings use different data!):

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

# Statistical Transformations:

Statistical transformations, in this context, refer to the process by which several geometric objects in ggplot2 are calculated.

Many graphs, like scatterplots, plot the raw values of your dataset. Other graphs, like bar charts, calculate new values to plot:

* Bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.

* Smoothers (like geom_smooth) fit a model to your data and then plot predictions from the model.

* Boxplots compute a robust summary of the distribution and then display a specially formatted box.

For this section we will use the dataset Diamonds, pre-loaded in base r
Let's take a look at  it:

```{r}
head(diamonds)
```

```{r}
str(diamonds)
```

```{r}
?diamonds
```


Let's take a look at a simple Bar Chart:

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

This barchart first counts the diamonds under each cut type, and once it does that, it proceeds to graphing the data. This counting procedure by a particular parameter is referred to as binning when it comes to bar charts. 

The statistical transformation in the data is referred to as a stat (in this case "count"). We can sometimes graph the same barchart by using a geom or its default stat. For instance, see how the code below yields the same graph as the one above:

```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

This works because every geom has a default stat; and every stat has a default geom. This means that you can typically use geoms without worrying about the underlying statistical transformation. 

**Overriding the stats within cetain Geoms:**

You can always override the default stat in a geom that works with stats. 

In the case of bar charts you might want to graph your data with the height of the rows that are already present in the data (without transforming your data using the count function). If this is the case, you will have to override the count stat with theidentity stat. 

You might also want to override the stats of other geometrical objects that work with statistical transformations. 

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

It is also useful to override the default main variable type from absolute numbers to another number format / transformation. For example, you might want to display a bar chart of a proportion, rather than count.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

Finally, you might want to draw greater attention to the statistical transformation in your code. For example, you might use stat_summary(), which summarises the y values for each unique x value, to draw attention to the summary that you're computing. In this case you wil not use a geom, but instead you will use a stat_summary parameter within GGPLOT2:

```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

ggplot2 provides over 20 stats for you to use. Each stat is a function, so you can get help in the usual way, e.g. ?stat_bin. To see a complete list of stats, try the ggplot2 cheatsheet.

# Position Adjustments

In this section we will cover certain aesthetics parameters for geometric objects that use statistical transformation, and focus mostly on position adjustments which modify how one views the data and helps performs actions such as stacking charts. 

To color a bar chart, it is best to use the "fill" aesthetic, rather than the color aesthetics that we used for charts that did not use statistical transformations:

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

If you map the fill aesthetic to another variable, such as clarity, then you automatically get a stacked chart, where each rectangle represents the combination of cut and clarity. This introduces more information to our chart. 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

The stacking within this chart is performed by the default "Position" adjustment within the geometric object. Nonetheless if you want to perform the stacking in a different way, there are some position adjustments that might help. 

**Position Adjustments** 

- position = "identity" : Will place each object exactly where it falls in the context of the graph. This is not very useful for bars, because it overlaps them. We can see this in the chart below by using some aesthetics that help us visualize this issue (NOTICE THE OVERLAPPING!)

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```

The identity position adjustment is more useful for 2d geoms, like points, where it is the default.

- position = "fill" : Works like stacking, but makes each set of stacked bars the same height. This makes it easier to compare proportions across groups. 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

- position = "dodge":  places overlapping objects directly beside one another. This makes it easier to compare individual values. This fixes the problem created by position = "identity" for bar charts

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

## Position Adjustments for Scatterplots

You can also use position adjustments with scatterplots. One useful way to use them is to correct for the data that is sometimes ommitted by a scatterplot when there are overlapping points (this is called overplotting). This makes it hard to determine where the mass of the data is. 

**To correct for this problem we can use the position adjustment "jitter"**. This adds a small amount of random noise to each point and shows us an approximation of where the mass of the data is.This spreads the points out because no two points are likely to receive the same amount of random noise.

Recall our first scatterplot. Did you notice that the plot displays only 126 points, even though there are 234 observations in the dataset?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

This is how this chart looks when using our position adjustment jitter:

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x = displ, y = hwy),position="jitter")
```

Because this is such a useful operation, ggplot2 comes with a shorthand for geom_point(position = "jitter"): geom_jitter().

To learn more about a position adjustment, look up the help page associated with each adjustment: ?position_dodge, ?position_fill, ?position_identity, ?position_jitter, and ?position_stack.

# Coordinate Systems

The default coordinate system is the Cartesian coordinate system where the x and y positions act independently to determine the location of each point. There are a number of other coordinate systems that are occasionally helpful.

Here I will only mention one coordinate systems adjustment: **coord_flip()**  

- coord_flip() switches the x and y axes. This is useful (for example), if you want horizontal boxplots. It's also useful for long labels: it's hard to get them to fit without overlapping on the x-axis.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

#The Layered Grammar of Graphics --> Wrapping up

In the previous sections we can see how versatile ggplot2 is (it allows us to do multiple charts, change their aesthetics, etc). 
You can actually do almost any type of plot with ggplot2! And you can think of ggplot as a layered interface that allows you to always build on top of it. 

To see this, let's add position adjustments, stats, coordinate systems, and faceting to our code template:

**GGPLOT TEMPLATE**

```{r, eval=FALSE}
  ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```
  
This template (a formal system for building plots) has 7 paremeters, although in practice you seldom use all seven, because ggplot2 will provide useful defaults for everything except the data, the mappings, and the geom function.

The grammar of graphics is based on the insight that you can uniquely describe any plot as a combination of **a dataset, a geom (type of graph), a set of mappings (variables to plot, aesthetics), a stat (statistical transformation), a position adjustment (i.e. stacking), a coordinate system (visualization direction), and a faceting scheme (split the graph into subplots).**


--------- [end of ddocument]


